An Introduction to Rocker: Docker Containers for R

### 2017-06-18 manuscript.Rmd
WIP

@@

1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
2. These Dockerfiles always install the most recent available software

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users.

### debian-stable stack

This stack emphasizes stability and reproducibility of the Docker build.

>>>>

## The debian:testing base

@@

1. These Dockerfiles are easy to develop and extend because almost all software can be installed through the package manager
2. These Dockerfiles always install the most recent available software
3. These Dockerfiles can almost always build relatively quickly

@@

1. These Dockerfiles require occassional minor maintenance to ensure successful builds
2. The resulting images are inherently dynamic: rebuilding the same Dockerfile months or years apart will generate images with significantly different versions of software installed.
3. The use of apt pinning may be unfamiliar to some users.
4. `testing` and `unstable` do not recieve the same attention to security updates.

### debian-stable base

This stack emphasizes stability and reproducibility of the Docker build.  The key feature of this stack is the ability to run older versions of R, along with the contemporaneous versions of R packages.  A user specifies the version desired using an image tag, e.g. 
`rocker/r-ver:3.3.1` will refer to an image with R version 3.3.1 installed.  Omitting the tag is equivalent to using the tag `latest`, which, as the name implies, will always point to an image using the current R release.  Thus, users wanting to create downstream Dockerfiles which are based on the current release at time but will continue to reconstruct the same environment in the future after newer R versions are released should explicilty include the corresponding version tag, e.g. `rocker/r-ver:3.4.0` at the time of writing, and not the `latest` tag.  Users can also run the current development version of R using the tag `devel`, which is built nightly from R-devel sources.

To facilitate installation of only contemporaneous versions of R packages on these images, the default CRAN mirror from which to install R packages is fixed to a snapshot of CRAN corresponding to the last date for which that version of R was the latest release. These snapshots are provided by the MRAN archive created by Revolution Analytics (now aquired by Microsoft), which archives daily snapshots of all of CRAN, from which a user can install packages with the usual `install.packages()` function  [@MRAN]. Users can always override this default by passing any current CRAN repository explicitly.  Unlike CRAN, Bioconductor only updates its repositories through bi-annual releases aligned to R's Spring release schedule, Bioconductor packages be installed in the usual way using `bioclite`, which automatically selects the Bioconductor release corresponding to the version of R in use.

These version tags are propegated throughout this stack: for instance, `rocker/tidyverse:devel` will provide the currently released versions of the R packages in the `tidyverse` [@tidyverse] installed on the nightly build of R-devel.  Developers building packages on this stack are encouraged to tag their images accordingly as well.  Table 3 indicates which versions of R are currently available in the stack, going back to `3.1.0`.  While older versions may be added to the stack at a later date, we note that the MRAN snapshots began in starting 2014-09-17 and thus go back only to R `3.1` era.

In this stack, the desired version of R is always built directly from source.  Compilers and dependencies are still installed from the stable `apt` repositories, and thus lag behind the more recent versions found in the `testing` stack.  Dependencies needed to compile R that are not required at runtime are removed once R is installed, keeping the base images light-weight for faster download times.  While most system dependencies required by common R packages can still be installed from the `apt` repositories, occassionally a more recent version must be compiled from source (e.g. the Gibbs Sampling library, JAGS, and the geospatial toolkit GDAL, must both be compiled from source on images based on the Debian 8.0 release, codenamed `debian:jessie`).  In this stack, users should never install R packages using `apt`, since this will install a second, older version of R from the Debian repositories and a dated version of the R package.  These

### Images

The image `rocker/r-ver` is the functional equivalent of `r-base`, though slightly pared down.  Currently `r-ver` has tags for all minor R versions back to `3.1.0` (as far back as MRAN snapshots of CRAN are avialable); see [all tags](https://hub.docker.com/r/rocker/r-ver/tags). Users can attempt to build other versions from the `r-ver` Dockerfile by specifying `--build-arg R_VERSION=<VERSION>`, though this is unlikely to work with very old R versions.  Likewise, all non-current images are automatically pinned to a MRAN snapshot of CRAN from the last day that version was current.  Users can set a custom snapshot date for the `r-ver` images by using `--build-arg BUILD_DATE=<DATE>` when building that Dockerfile.

The image `rocker/tidyverse` is the functional equivalent of `rocker/hadleyverse`, minus the LaTeX part which is moved into a downstream `rocker/verse` in order to keep the `tidyverse` image more compact.  Now that [tidyverse](https://cran.r-project.org/web/packages/tidyverse/index.html) is a thing, the term "hadleyverse" is on the way out.

The `rocker/rstudio` image builds with RStudio `v1.0.44` by default. This can be customized by specifying the desired version in `--build-arg RSTUDIO_VERSION=<VERSION>` if building locally from its Dockerfile. *Note that:* `rocker/rstudio:latest` is the only image name in this stack that already existed.  For consistency, `rocker/rstudio:latest` now builds on `r-ver:latest` instead of `r-base:latest`.  This change may break downstream images, which should either build explicitly on the `debian:testing` stack by using the `testing` tag (`rocker/rstudio:testing`) or be patched to support this change.

image            |  Dockerfiles  
---------------- | -------------
[r-ver](https://hub.docker.com/r/rocker/r-ver/tags)         | [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/3.3.2/Dockerfile)  [![](https://img.shields.io/badge/Dockerfile-...-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/)  [![](https://img.shields.io/badge/Dockerfile-v3.1.0-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/r-ver/3.1.0/Dockerfile) 
[rstudio](https://hub.docker.com/r/rocker/rstudio/tags)     |  [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/rstudio/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker/blob/master/rstudio/3.3.2/Dockerfile)  [![](https://img.shields.io/badge/Dockerfile-v3.3.1-blue.svg)](https://github.com/rocker-org/rocker/blob/master/rstudio/3.3.1/Dockerfile) 
[tidyverse](https://hub.docker.com/r/rocker/tidyverse/tags) | [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/tidyverse/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/tidyverse/3.3.2/Dockerfile)   [![](https://img.shields.io/badge/Dockerfile-v3.3.1-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/tidyverse/3.3.1/Dockerfile) 
[verse](https://hub.docker.com/r/rocker/verse/tags)         |  [![](https://img.shields.io/badge/Dockerfile-latest-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/verse/Dockerfile) [![](https://img.shields.io/badge/Dockerfile-v3.3.2-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/verse/3.3.2/Dockerfile)  [![](https://img.shields.io/badge/Dockerfile-v3.3.1-blue.svg)](https://github.com/rocker-org/rocker-versioned/blob/master/verse/3.3.1/Dockerfile)

## Maintenance and Updates

These images are actively maintained.  This means that while an effort is made to preserve the general function of these images over time, both these Dockerfiles and the resulting images are subject to some change over time.  In particular:

- Images are regularly re-built on Docker Hub whenever their base image changes, starting with changes to `debian` Docker image.  This is the rough equivalent of running `apt-get upgrade` on `debian`, since all `apt-get` commands are re-run and will pull in the most current sources.  This allows the images to receive security updates to any packages installed from the `debian` repositories, but will not in general change the versions of any software and is very unlikely to break anything.

- The Dockerfiles themselves are subject to change, to improve performance, ease of use, readability, or other concerns raised in the issues.  These changes should also not alter the general behavior of R or R packages on the image.  These changes can be seen in the git history.  The [rocker-versioned](https://github.com/rocker-org/rocker-versioned) repo will use its own semantic version tagging to indicate changes to this repository, with snapshots from these tags archived on Zenodo.

## License ##

The Dockerfiles in this repository are licensed under the GPL 2 or later.
